\begin{abstract}
Snapshot-backed sandboxes make branching cheap while leaving evidence dependence unchanged.
Branches can reuse a model, prompt, repository, tests, observations, or execution ancestor, so
counting outputs can amplify one repeated error into high-confidence consensus. We introduce an
\emph{evidence-aware reduction contract}: each worker reports an estimate, estimated
information, evidence identifiers, fork lineage, and execution metadata. For independent workers
estimating one common parameter, we use standard inverse-information pooling in its Gaussian/Wald
form. The fixed-dimensional numeric summary can merge in any tree order; evidence IDs and lineage
follow separate rules. The residual $\Delta$ measures disagreement, becomes Cochran's $Q$ in the
scalar inverse-variance case, and appears in the product integral. A reference implementation
validates serialized records, rejects repeated nonempty evidence identifiers, carries evidence
and lineage through tree reduction, and uses Cholesky-based numerical linear algebra. Unit tests
and seeded synthetic checks exercise the algebra, unequal information, and forged precision; one
four-worker named-snapshot trace exercises the end-to-end path. Platform logs document the
exercised execution paths. A central open systems challenge is to turn evidence identity
and fork lineage into a dependence model for correlated and adaptively selected AI branches.
Artifact: \href{https://github.com/zozo123/boltzmann-mapreduce}{github.com/zozo123/boltzmann-mapreduce}.
\end{abstract}

\section{Introduction}
\label{sec:intro}

\textbf{Execution fan-out and evidence fan-out are different resources.} A snapshot runtime can
restore one prepared environment into many isolated workers, but shared models, prompts, tests,
data, repositories, and ancestors can induce common-mode error. If $K$ scalar estimates have
equal marginal variance $\sigma^2$ and exchangeable pairwise correlation $\rho$, then
\begin{equation}
\operatorname{Var}(\bar\theta)=
\sigma^2\!\left[\rho+\frac{1-\rho}{K}\right],
\label{eq:correlation-floor}
\end{equation}
whereas an independence calculation reports $\sigma^2/K$. For fixed $\rho>0$, variance has
the floor $\rho\sigma^2$ even as the number of branches grows. Execution isolation constrains
runtime state; $\rho$ instead depends on the evidence sources and branch history.

MapReduce gained leverage from a small interface that let the runtime manage partitioning,
placement, retries, and communication~\cite{dean2004mapreduce}. Modern evaluation and agent
pipelines often retain the fan-out but reduce each branch to a scalar or vote. Such outputs erase
four facts a reducer needs: uncertainty, information strength, evidence identity, and inherited
dependencies. Equal voting ignores the first two; repeated evidence and shared ancestry can
manufacture false precision from the latter two. Snapshot and serverless systems make this
failure mode operationally important because branching prepared state is increasingly
convenient~\cite{agache2020firecracker,ustiugov2021reap,du2020catalyzer}.

We ask a systems question: \emph{what should a worker return so that a runtime can reduce its
result without erasing the evidence behind it?} Our answer is a structured worker record and a
mergeable canonical state. The implementation covers the independent common-target case,
rejects literal reuse of caller-supplied evidence identifiers, and carries lineage into the
final result. Turning that lineage into a dependence estimate is the next research step.

\noindent\textbf{Contributions.}
\begin{itemize}
  \item \textbf{Interface.} An evidence-carrying worker record and a fixed-dimensional Gaussian
  numeric state for flat, streaming, and tree reduction, with provenance aggregation and exact
  duplicate-ID rejection.
  \item \textbf{Artifact.} A reference implementation with record validation, closed-form
  product-integral evaluation, Cholesky solves, deterministic statistical checks, and a
  snapshot-backed integration trace.
  \item \textbf{Scope and agenda.} A precise separation between execution isolation and
  independent evidence, exposing correlation inherited through fork DAGs, confidence
  calibration, adaptive selection, and forged precision as reducer requirements.
\end{itemize}

\section{Evidence-Aware Worker Results}
\label{sec:contract}

\input{fig_contract}

\paragraph{Wire record.}
Worker $k$ emits
\begin{equation}
r_k=(\hat\theta_k,J_k,n_k,\mathcal E_k,\mathcal L_k,m_k),
\label{eq:contract}
\end{equation}
where $\hat\theta_k\in\mathbb R^p$ is an estimate, $J_k$ is estimated
per-observation information, $n_k$ is a literal sample count, $\mathcal E_k$ contains evidence
identifiers, $\mathcal L_k$ records fork lineage, and $m_k$ stores execution metadata. The
Python dataclass has a JSON-compatible serialized form. At ingestion, the implementation rejects
non-finite estimates, non-positive integer counts, malformed provenance, inconsistent dimensions,
and non-symmetric or non-positive-definite $J_k$. The current contract requires positive
definiteness; the algebra also admits semidefinite local factors when their aggregate precision
is full rank.

The invariant quantity is total precision
\begin{equation}
P_k=n_kJ_k,\qquad q_k=P_k\hat\theta_k,\qquad
c_k=\hat\theta_k^TP_k\hat\theta_k.
\label{eq:canonical}
\end{equation}
The decomposition $P_k=n_kJ_k$ is useful when $n_k$ is an actual observation count; workers with
bootstrap, clustered, weighted, or verifier-derived uncertainty may instead require a direct
precision field and a separate evidence-mass convention.

\paragraph{Trust semantics.}
Evidence identifiers and lineage are optional caller-supplied declarations whose reliability
rests on the caller. The reducer compares nonempty $\mathcal E_k$ sets and rejects exact overlap
by default. This check covers literal ID reuse; empty identifiers, relabeling the same data, and
shared noise fall outside it. The record carries dimensions but no validated estimand, unit,
model-version, or coordinate identifier. A production runtime must bind these fields to
authenticated data and model manifests before treating equal dimensions as a common target.

\paragraph{Numeric merge and provenance.}
Independent summaries add
\begin{equation}
(P,q,c,N)=\bigoplus_k(P_k,q_k,c_k,n_k).
\label{eq:merge}
\end{equation}
In exact arithmetic the numeric state is associative and commutative, enabling arbitrary tree
reduction. The scalar $c$ must survive intermediate merges: re-emitting only
$(\hat\theta,J,N)$ would erase disagreement accumulated below the tree. Provenance follows its
own semantics. On merge, evidence IDs are canonicalized as a set; lineage uses order-preserving
deduplication, so the full serialized summary is order-sensitive. The final result returns both
provenance fields. Numeric state costs $O(p^2)$, whereas an exact
provenance ledger can grow as $O(|\mathcal E|+|\mathcal L|)$.

\paragraph{Relationship to sufficient-statistic reduction.}
Model-specific sufficient statistics remain preferable when available: a mean is recovered from
$(\sum_i x_i,n)$ and ordinary least squares from $(X^TX,X^Ty)$. The contract targets one-shot
settings where workers expose estimates and calibrated uncertainty but raw data or
model-specific sufficient statistics are unavailable or undesirable to transmit.

\paragraph{Execution backends.}
The same record works across processes, containers, and microVMs. The local backend uses
Python's platform-dependent process pool. The documented islo workflow prepares and saves a
named environment before the normal backend restores one sandbox per shard. This separation
lets an execution planner choose placement and tree shape without changing the statistical
contract.

\paragraph{Retries and speculative execution.}
An evidence token identifies an evaluation independently of the process that executes it. A
retry after failure and a speculative duplicate should therefore reuse the original evidence
token; accepting both would turn fault tolerance into double counting. The current overlap guard
raises an error when it sees a repeated token. A production keyed admission layer should retain
one result for retries or speculative duplicates. A branch that draws a genuinely new
sample receives a new token while retaining its parent lineage. Production tokens should bind
the input partition, evaluator, model and prompt versions, and selection context through
authenticated structured records.

\section{Associative Gaussian Reduction}
\label{sec:theory}

\paragraph{Statistical scope.}
Confidence distributions and inverse-information methods combine evidence from independent
sources~\cite{singh2005combining,xie2011meta,xie2013confidence}; Rao-CD places
estimating-function inference in a MapReduce setting~\cite{zhou2017raocd}. We use their
Gaussian/Wald common-effect special case. Assume disjoint evidence with independent estimation
noise, a common parameter $\theta_0$, fixed $p$ and $K$, $\min_k n_k\to\infty$, and calibrated
positive-definite precision estimates. Write $V_k$ for worker $k$'s per-observation asymptotic
covariance and suppose
\begin{equation}
n_k^{1/2}(\hat\theta_k-\theta_0)
\overset{d}{\longrightarrow}N_p(0,V_k),\qquad
J_k\overset{p}{\longrightarrow}V_k^{-1}.
\label{eq:an}
\end{equation}
For efficient likelihood workers, local asymptotic normality is one route to this display; for
estimating-equation workers, $V_k$ is a sandwich covariance and $V_k^{-1}$ is Godambe
precision~\cite{godambe1960estimating}. The leading Wald confidence kernel is
\begin{equation}
g_k(\theta)=\exp\!\left\{-\tfrac12
(\theta-\hat\theta_k)^TP_k(\theta-\hat\theta_k)\right\}.
\label{eq:factor}
\end{equation}
It is exact when the local estimate is Gaussian with fixed known covariance. With plug-in
covariance or non-Gaussian errors, it is a first-order asymptotic approximation.

\paragraph{Pooled center and heterogeneity.}
Let $P=\sum_kP_k$, $q=\sum_kq_k$, and $c=\sum_kc_k$. Completing the square gives
\begin{equation}
\hat\theta=P^{-1}q,\qquad \Sigma_g=P^{-1},
\label{eq:pool}
\end{equation}
where $\Sigma_g$ is exactly the covariance of the normalized product kernel and, under
Eq.~\eqref{eq:an}, a first-order estimate of the sampling covariance of $\hat\theta$. The
residual quadratic is
\begin{equation}
\begin{aligned}
\Delta
&=\sum_k(\hat\theta_k-\hat\theta)^TP_k
       (\hat\theta_k-\hat\theta)\\
&=c-q^TP^{-1}q\ge0
\quad\text{in exact arithmetic}.
\end{aligned}
\label{eq:heterogeneity}
\end{equation}
For $p=1$, $\Delta$ is Cochran's $Q$; the vector expression is the corresponding Gaussian
weighted-residual diagnostic with direct multivariate meta-analytic analogues
~\cite{cochran1954combination,jackson2012heterogeneity}. Under independent
$N_p(\theta_0,P_k^{-1})$ estimates with fixed known $P_k$,
$\Delta\sim\chi^2_{p(K-1)}$; plug-in precision makes this calibration asymptotic. Small values
remain compatible with dependence or duplicated evidence. Comparisons across different $K$
must account for degrees of freedom.

For the unit-height kernels in Eq.~\eqref{eq:factor}, the partition integral is
\begin{equation}
Z_g=(2\pi)^{p/2}|P|^{-1/2}\exp(-\Delta/2).
\label{eq:normalizer}
\end{equation}
Thus $\Delta/2$ is both the minimized summed quadratic energy and the heterogeneity contribution
to $-\log Z_g$. The full integral also depends on $|P|$ and the integration measure. Its
magnitude alone cannot be read as a conflict score or model evidence. Normalized local kernels
require their own Gaussian constants.

\paragraph{Numerics and naming.}
The implementation obtains $\hat\theta$, $\Sigma_g$, and $\log|P|$ from a Cholesky factor,
avoiding explicit matrix inversion. It evaluates Eq.~\eqref{eq:normalizer} in closed form, clips a
negative computed $\Delta$ only within a scale-aware roundoff tolerance, and rejects larger
violations of the canonical invariant; the subtraction can otherwise suffer cancellation at
large offsets. Writing
\begin{equation}
E_k(\theta)=\tfrac12(\theta-\hat\theta_k)^TJ_k(\theta-\hat\theta_k),
\qquad \beta_k=n_k
\label{eq:gibbs}
\end{equation}
makes $g_k=\exp\{-\beta_kE_k\}$ an unnormalized Gibbs kernel. This is a naming convention:
$n_k$ can be absorbed into $J_k$, and $P_k$ is the invariant. A global
$\beta=N=\sum_kn_k$ with weighted-average energy is the same product. Bounded local sample sizes
leave the Gaussian approximation unjustified, and accumulated local bias can dominate when $K$
grows~\cite{zhang2013oneshot,zhou2017raocd}.

\paragraph{Limits of reported precision.}
Holding reported precisions fixed, worker $o$ has local influence
\begin{equation}
\frac{\partial\hat\theta}{\partial\hat\theta_o}=P^{-1}P_o.
\label{eq:influence}
\end{equation}
Low relative precision in a direction limits leverage there, but distance leaves the derivative
unchanged; fabricated high precision creates another way to control the pooled estimate. The
artifact's stress-test clip caps determinant-based precision volume and applies a coordinate-wise median/MAD
redescending location weight. A determinant cap leaves the largest directional eigenvalue
unbounded, and the location rule depends on the coordinate system. This heuristic instruments
one failure mode without providing a Byzantine robustness guarantee.

\begin{table*}[t]
\centering\scriptsize
\caption{Platform-specific batched traces. Daytona and Tensorlake percentiles cover sandbox
creation; their total wall time covers the complete create--run--delete batch. The islo
percentiles cover restore--run--capture round trips.}
\label{tab:platform}
\begin{tabularx}{\textwidth}{@{}l l >{\raggedright\arraybackslash}X r r r r r r@{}}
\toprule
Platform & State source & Measured operation & Client conc. & Batch & Success & $p_{50}$ & $p_{95}$ & Total wall \\
\midrule
Daytona & default environment & create & 8 & 1024 & 1024/1024 & 0.20\,s & 1.12\,s & 190.3\,s \\
Tensorlake & default environment & create & 8 & 256 & 256/256 & 3.44\,s & 9.00\,s & 387.3\,s \\
islo & named 141\,MB snapshot & restore + run + capture & 12 & 256 & 255/256 & 6.87\,s & 9.04\,s & 240.3\,s \\
\bottomrule
\end{tabularx}
\end{table*}

\section{Evaluation of the Reduction Contract}
\label{sec:eval}

The evaluation separates unit-tested identities, seeded behavioral checks, and execution traces.
All statistical data are synthetic. Table~\ref{tab:checks} states the conclusion supported by
each check. The point gaps summarize these runs; estimating statistical efficiency would require
a broader repeated-sampling study.

\begin{table}[t]
\centering\footnotesize
\caption{Artifact checks and their supported claims. The logistic values are mean $\pm$ SD over
eight seeds.}
\label{tab:checks}
\begin{tabularx}{\columnwidth}{@{}>{\raggedright\arraybackslash}p{0.23\columnwidth}>{\raggedright\arraybackslash}p{0.35\columnwidth}X@{}}
\toprule
Check & Observation & Supported conclusion \\
\midrule
Canonical algebra & Flat/tree and closed-form results agree & Code matches the stated Gaussian formulas \\
Evidence overlap & Repeated nonempty ID raises an error & Exact labeled reuse is blocked \\
Unequal-size logistic & pool gap $0.0083\!\pm\!0.0042$; equal average $0.177\!\pm\!0.090$ & Information pooling improves on unweighted averaging in this check \\
Forged precision & attacked $17.0004$; clipped $4.9566$ & Vulnerability and behavior of one heuristic \\
Snapshot integration & four workers complete in $6.70$\,s & Named-snapshot path executes end to end \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Algebra and local estimators.}
Twelve homoscedastic mean shards of size $107$--$355$ produce $4.9542$ with a nominal plug-in
$95\%$ Wald interval $[4.8980,5.0104]$; the full-sample mean is $4.9546$. The reducer agrees
with closed-form inverse-variance pooling to floating-point precision, and flat and tree
reductions agree. This checks algebra; $(\sum x,n)$ would recover the sample mean exactly. In
the linear-regression check, the pooled coefficient has norm gap $0.033$ to centralized OLS.
Each worker estimates its own residual variance, whereas exact aggregation of $X^TX$ and $X^Ty$
uses a different centralized variance treatment.

\paragraph{Nonlinear and adversarial checks.}
Five IID logistic shards have sizes $\{60,120,400,2000,5000\}$. Across eight fixed seeds,
information pooling is $0.0083\pm0.0042$ from the centralized MLE, while equal coefficient
averaging is $0.177\pm0.090$ away. We use unweighted averaging as a simple baseline; a broader
study would also compare sample-size weighting, Rao-CD, surrogate likelihood, and distributed
Newton. In the forged-precision check, one scalar worker reports a
distant location from a
2,000-point shard and inflates its reported precision by a further $50\times$; unprotected
pooling moves to $17.0004$, and the stress
clip returns $4.9566$. This check exposes the vulnerability and illustrates one narrow response.

\paragraph{Execution paths.}
One integration trace starts from a named $141$\,MB islo snapshot and restores four sandboxes.
Their seed-pinned workers return pooled mean $4.9422$ versus full-sample $4.9450$; four
concurrent restore--run--capture round trips complete in $6.70$\,s. This validates the
snapshot-to-worker-to-reducer path with client-observed timing.

Table~\ref{tab:platform} records the operation exercised on each platform. Higher requested
concurrency reached validation or capacity limits for Daytona at 32 and islo at 16 and 32.

\section{Assumptions and Open Problems}
\label{sec:agenda}

\paragraph{Dependence.}
Literal ID deduplication catches exact declared reuse. A lineage union records common ancestors
without encoding a fork DAG or mapping ancestry to covariance. A correlation-aware
reducer should combine factors conditionally on shared observations, evaluators, model state,
and selection history. Equation~\eqref{eq:correlation-floor} makes this a central statistical
problem.

\paragraph{From lineage to a fork-DAG reducer.}
A useful next record would name parent branch IDs, immutable evidence manifests, model/prompt/tool
hashes, evaluator and test versions, and branch-selection events. The runtime could then map
known overlap to shared latent factors, covariance blocks, or conservative evidence groups
before applying a generalized reducer. Crucially, the interface must permit abstention: when the
runtime cannot justify a dependence model, it should report unresolved correlation and withhold
a narrower interval. The current lineage union records the history while leaving this modeling
step open.

\paragraph{Different local targets.}
The common-$\theta_0$ assumption defines a common-effect model. Non-IID observations can still
support pooling when every worker identifies the same parameter and reports valid uncertainty.
If workers target different estimands, $P^{-1}$ omits between-worker variation and
can be severely overconfident. The system must first name the target and adopt an explicit
heterogeneity model, such as site effects, meta-regression, or a random-effects hierarchy.

\paragraph{Confidence construction and adaptive selection.}
A model's verbal confidence requires empirical calibration before it can serve as a standard
error; calibration depends on the model, elicitation, task, and deployment distribution. Agent
precision should be derived from external evidence such as held-out tests, repeated trials,
independent verifiers, or an empirically calibrated outcome model. If the same evidence selects
surviving branches and constructs their factors, winner's-curse bias can result; the ledger must
record that reuse.

\paragraph{Untrusted workers and systems evidence.}
Robust aggregation requires a threat model over locations, covariance reports, collusion, and
dimension. Controlled systems comparisons likewise require identical images and workloads,
concurrent fan-out, tail latency, memory sharing, failures, cost, and statistical utility per
unit time. The present artifact supplies interfaces and traces on which those studies can be
built.

\section{Related Work}
\label{sec:related}

The Gaussian reducer follows established confidence-distribution and common-effect
meta-analysis~\cite{singh2005combining,xie2011meta,cochran1954combination}. Rao-CD and
distributed GLM inference provide closely related estimating-equation constructions
~\cite{zhou2017raocd,tang2020distributed}. One-shot averaging and surrogate likelihood explore
neighboring communication--statistics tradeoffs~\cite{zhang2013oneshot,jordan2019csl}. We add a
runtime contract that makes the independence conditions visible before reduction.

Self-consistency marginalizes answers from diverse reasoning paths, while multiagent debate lets
models exchange critiques before converging~\cite{wang2023selfconsistency,du2024multiagent}.
These methods exploit redundant generations for answer selection while leaving evidence
dependence unstated. Our contract records provenance, and its Gaussian reducer combines
uncertainty when the independence assumptions hold.

MapReduce established the programmable dataflow contract~\cite{dean2004mapreduce}; Firecracker,
Catalyzer, REAP, and Faasm study lightweight isolation, startup, and reusable state
~\cite{agache2020firecracker,du2020catalyzer,ustiugov2021reap,shillaker2020faasm}. These systems
make branching practical; our work connects that capability to explicit evidence accounting at
reduction time. Robust distributed
learning methods such as Krum and FLAME start from explicit adversary models
~\cite{blanchard2017krum,nguyen2022flame}. The included stress clip is a diagnostic heuristic
without a Byzantine-robustness guarantee.

\section{Conclusion}
\label{sec:conclusion}

Cheap forks reduce execution cost while leaving the amount of independent evidence unchanged. The
contract developed here makes an estimate, uncertainty, evidence identity, and lineage explicit,
then reduces independent Gaussian summaries through an associative numeric state while retaining
provenance. The reference artifact validates this path and reports residual heterogeneity
explicitly. The next systems step is a reducer over the fork DAG: one that can
explain why two branches count as separate evidence, model the dependencies they inherited, and
remain calibrated after selection. Faster fan-out is useful; evidence-aware fan-in keeps the
assumptions and unresolved dependencies visible.
